---
title: "FA4 DSC1105"
author: "Lindsay Faith Bazar"
date: "March 07, 2025"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## 1. Mortality vs.Temperature

```{r load-data}

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})

data <- read.csv("mortality_by_latitude.csv")

colnames(data)
summary(data)

```

### Plot

```{r plot-mortality}
p1 <- ggplot(data, aes(x = temperature, y = mortality_index)) +
  geom_point() +
  geom_smooth(method = "loess") +
  theme_minimal() +
  ggtitle("Mortality Index vs. Temperature")

print(p1)
```

The plot is hollow up because the curve dips a little at first and then rises as the temperature increases. It has a slight U-shape, which means the relationship between mortality index and temperature curves upward rather than bending downward.

### Transformation 

```{r transform-mortality}

data <- data %>% mutate(log_mortality_index = log(mortality_index))

p2 <- ggplot(data, aes(x = temperature, y = log_mortality_index)) +
  geom_point() +
  geom_smooth(method = "loess") +
  theme_minimal() +
  ggtitle("Log Mortality Index vs. Temperature")

print(p2)

```

### Residuals

```{r residuals-mortality}

model <- lm(log_mortality_index ~ temperature, data = data)
data$residuals <- resid(model)

p3 <- ggplot(data, aes(x = temperature, y = residuals)) +
  geom_point() +
  geom_smooth(method = "loess") +
  theme_minimal() +
  ggtitle("Residual of the Transformed Model")

print(p3)
```

## 2. Diamond Dataset


```{r load-diamonds}

data(diamonds)
diamonds_subset <- diamonds %>% filter(carat < 2.5)
```

### Span

```{r span }

ggplot(diamonds, aes(x = carat, y = log(price))) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", span = 0.2, color = "red", size = 1, se = FALSE) +  
  geom_smooth(method = "loess", span = 0.5, color = "blue", size = 1, se = FALSE) + 
  geom_smooth(method = "loess", span = 0.8, color = "green", size = 1, se = FALSE) + 
  ggtitle("LOESS Smoothing with Different Span Values")

```

The span value controls how smooth or wiggly the curve is. The red line uses a small span (0.2), so it responds to even the smallest change in the data, making it appear more wiggly since it captures the smallest details.  The green line has the largest span (0.8), so it smooths out the ups and downs, showing only the general trend, which miss some minor details. The blue line (0.5) seems to capture the good balance of the other two, it gives the overall trend without being too smooth or too wiggly.

So if we wish to capture minor details, we use a smaller span (0.2). If we just need the overall trend, we use the larger span (0.8). And if we want both the details and the general trend, we use the middle span (0.5)

## 3. Model Comparison

```{r loess fit}
loess_fit <- loess(price ~ carat, data = diamonds)
diamonds$loess_resid <- residuals(loess_fit)

ggplot(diamonds, aes(x = carat, y = loess_resid)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals of LOESS Model",
       x = "Carat", y = "Residuals") +
  theme_minimal()
```

```{r Fit Polynomial + Step Function Model}
poly_model <- lm(price ~ poly(carat, 3) + cut, data = diamonds)
diamonds$poly_resid <- residuals(poly_model)

ggplot(diamonds, aes(x = carat, y = poly_resid)) +
  geom_point(color = "red", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals of Polynomial + Step Function Model",
       x = "Carat", y = "Residuals") +
  theme_minimal()
```

### Side by side

```{r ss}

diamonds_resid <- diamonds %>%
  select(carat, loess_resid, poly_resid) %>%
  pivot_longer(cols = c(loess_resid, poly_resid), names_to = "Model", values_to = "Residuals")

ggplot(diamonds_resid, aes(x = carat, y = Residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ Model, scales = "free_y") +
  labs(title = "Residuals Comparison: LOESS vs. Polynomial + Step Function",
       x = "Carat", y = "Residuals") +
  theme_minimal()

```

Residuals measure how far off a model’s predictions are from actual values. In the LOESS plot, residuals seems to be spread out evenly, with no clear pattern. While the polynomial + step function seems to cluster in certain areas. In this regard, LOESS is more faithful because it has no obvious structure and doesn’t leave a clear pattern.  