---
title: "DSC1105_FA3"
author: "Lindsay Faith Bazar"
date: "February 28, 2025"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(ggplot2)
library(tidyverse)
```

```{r}
data("diamonds")
```

## 1. Create a histogram on the diamonds dataset

```{r}
ggplot(diamonds, aes(x = carat)) + 
  layer(
    geom = "bar",
    stat = "bin",
    position = "stack",
    mapping = aes(y = after_stat(count)),
    params = list(bins = 30)
  )
```

## 2. Modify the histogram to use a different geom.

```{r}
ggplot(data = diamonds, aes(x = carat)) + 
  layer(
    geom = "line",
    stat = "bin",
    position = "identity",
    mapping = aes(y = after_stat(count)),
    params = list(binwidth = 0.1)         
  )
```

## 3. In your histogram, add an aesthetic mapping from one of the factor variables to the fill or color aesthetic.

```{r}
ggplot(diamonds, aes(x = carat)) + 
  layer(
    geom="bar",
    stat="bin",
    position="stack",
    mapping= aes(y = after_stat(count), fill=color),
    params = list(bins = 30)
  )
```

## 4. What is the default position adjustment for a histogram? Try changing the position adjustment in the histogram.

By default, histograms use position = "stack", meaning bars are placed on top of each other.

If we change it to position = "dodge", bars will now be then placed side by side.

```{r}
ggplot(diamonds, aes(x = carat, fill = color)) +
  layer(
    geom = "bar",
    stat = "bin",
    position = "dodge",
    mapping = aes(y = after_stat(count)),
    params = list(bins = 30)
  )
```