---
title: "SA2 DSC1105"
author: "Lindsay Faith Bazar"
date: "May 18, 2025"
output: html_document
---


```{r setup, include=FALSE}
options(warn = -1)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)         
library(nnet)        
library(pscl)
library(ggplot2)
library(multcomp)    
library(stats)
```

## **Introduction**

&emsp; &emsp; This study focuses on identifying social and demographic predictors of pregnancy outcomes among women in the United States. The data used in this study come from the 2022–2023 Female Respondent Public Use File of the National Survey of Family Growth (NSFG), a nationally representative survey conducted by the National Center for Health Statistics (NCHS). The survey collects detailed information about family life, reproductive health, marriage, contraception, and related topics from women aged 15–49 in the U.S.


&emsp; &emsp; The primary objective of this analysis is to examine how a woman’s background, education, marital status, and family structure relate to the total number of pregnancies a woman have. Using Poisson and negative binomial regression models, this study explores the associations between these predictors and pregnancy numbers. Additional statistical methods, including ordinal and multinomial logistic regression, chi-square tests, and data visualizations, are used to support a deeper understanding of patterns in the data.


```{r}
data <- read_csv("NSFG_2022_2023_FemRespPUFData.csv")
data
```

A subset of variables was selected for this analysis:


```{r echo=FALSE}
variable_table <- data.frame(
  Variable = c("PREGNUM", "FMARITAL", "HISPRACE2", "HIEDUC", "AGER", "EDUCMOM", "INTCTFAM"),
  Description = c(
    "Total number of pregnancies reported by the respondent (response variable).",
    "Current formal marital status of the respondent (e.g., married, divorced, never married).",
    "Respondent’s race and Hispanic origin, categorized as White, Black, or Other.",
    "Highest level of education completed by the respondent, treated as an ordered categorical variable.",
    "Age of the respondent at the time of the interview (continuous variable).",
    "Education level of the respondent’s mother or female caregiver, used as a proxy for childhood socioeconomic background.",
    "Indicates whether the respondent grew up in an intact family (i.e., living with both biological or adoptive parents)."
  )
)

knitr::kable(variable_table, format = "markdown")
```


## **Variable Selection** 

#### ***Dependent Variable ($Y$)***

**Count Variable**

`PREGNUM` - Total Number of Pregnancies

This is the main outcome of the analysis. We want to understand what factors are are related to how many times a woman has been pregnant. 

```{r}
summary(data$PREGNUM)
```

#### ***Independent Variable ($X_i$)***

**Categorical (nominal)**

`FMARITAL` - Marital Status


Whether someone is married, divorced, or never married can influences their chances of getting pregnant. For example, married women might be more likely to plan for children.

```{r}
summary(data$FMARITAL)
```

`HISPRACE2` - Race and Hispanic Origin

This can help see if pregnancy patterns are different across group, possibly due to things like culture or access to healthcare. 

```{r}
summary(data$FMARITAL)
```

**Ordered Categorical**

`HIEDUC` - Highest Educational Level

For example, women with more education might focus on school or work first and have fewer, later, or no plans of pregnancies.

```{r}
summary(data$HIEDUC)
```

**Demographics**

`AGER` - Age

The older someone is, the more time they've had to potentially have pregnancies.

```{r}
summary(data$AGER)
```

`EDUCMOM`- Mother's Education

Growing up with an educated mother might shape attitudes towards family planning.

```{r}
summary(data$EDUCMOM)
```

`INTCTFAM` - Grew Up in an Intact Family or Not

This looks at whether women lived with both parents growing up. Family structure might affect emotional development and future life choices, including when or whether to have children.

```{r}
summary(data$INTCTFAM)
```

## **Data Cleaning**

**Checking null values:**

```{r}
vars <- c("PREGNUM", "FMARITAL", "HIEDUC", "HISPRACE2", "AGER", "EDUCMOM", "INTCTFAM")

sapply(data[vars], function(x) sum(is.na(x)))
```

**Recoding categorical variables:**

```{r}

data <- data %>%
  mutate(
    FMARITAL = factor(FMARITAL,
                      levels = c(1, 2, 3, 4, 5),
                      labels = c("Married", "Widowed", "Divorced", "Separated", "Never Married")),
    HISPRACE2 = factor(HISPRACE2,
                       levels = c(1, 2, 3, 4),
                       labels = c("Hispanic", "White", "Black", "Other")),
    HIEDUC = factor(HIEDUC, ordered = TRUE), 
    EDUCMOM = factor(EDUCMOM, ordered = TRUE),
    INTCTFAM = factor(INTCTFAM, levels = c(1, 2), labels = c("Intact", "Not Intact"))
  )

```


## **Statistical Analyses**

#### **Poisson Regression**

```{r}
poisson_model <- glm(PREGNUM ~ FMARITAL + HISPRACE2 + HIEDUC + AGER + EDUCMOM + INTCTFAM,                  family = poisson(link = "log"),
                     data = data)
```

```{r}
summary(poisson_model)
```

To assess whether the Poisson regression model was appropriate, we checked for overdispersion, by examining the ratio of the residual deviance to its degrees of freedom.

```{r}
dispersion <- sum(residuals(poisson_model, type = "pearson")^2) / poisson_model$df.residual
dispersion 
```

The Poisson regression model showed evidence of overdispersion (dispersion statistic > 1).

To address this, a Negative Binomial model was fitted to account for the extra-Poisson variation.

```{r}
nb_model <- glm.nb(PREGNUM ~ FMARITAL + HISPRACE2 + HIEDUC + AGER + EDUCMOM + INTCTFAM,
                     data = data)
summary(nb_model)
```

Model fit was compared using the Akaike Information Criterion (AIC).

```{r}
AIC(poisson_model, nb_model)
```

The Negative Binomial model provided a better fit to the data, with a lower AIC (X) compared to the Poisson model (Y). Based on these results, the Negative Binomial model was considered the more appropriate model for inference.

**Summary of significant predictors:**

```{r}

coefs <- coef(nb_model)

IRRs <- exp(coefs)

percent_change <- (IRRs - 1) * 100

pvalues <- summary(nb_model)$coefficients[, "Pr(>|z|)"]

table <- data.frame(
  Estimate = coefs,
  IRR = IRRs,
  Percent_Change = round(percent_change, 2),
  P_Value = pvalues
)

significant_predictors <- table %>% filter(P_Value < 0.05)

significant_predictors
```


&emsp; The negative binomial regression model revealed several important factors associated with the number of pregnancies among women. Compared to married women, those who were never married had 58% fewer pregnancies (IRR = 0.42), and those who were divorced had 22% fewer pregnancies (IRR = 0.78). Widowed women also had fewer pregnancies (IRR = 0.73).

&emsp; In terms of race, Black women had 38% more pregnancies than Hispanic women (IRR = 1.38), while women classified as Other races had 15% fewer (IRR = 0.85). White women had slightly fewer pregnancies (IRR = 0.94).

&emsp; Education showed a clear pattern: as education increased, the number of pregnancies generally decreased. For example, the main trend (HIEDUC.L) showed that each step up in education level was associated with a 50% decrease in the number of pregnancies (IRR = 0.50).

&emsp; Age also had a strong positive effect: for each additional year of age, the expected number of pregnancies increased by about 5.6% (IRR = 1.06), as older women naturally have had more time to become pregnant.

&emsp; Looking at background factors, women whose mothers had some college education had 25% more pregnancies (IRR = 1.25), while those whose mothers had less than high school had slightly fewer (IRR = 0.88). Additionally, women who did not grow up in an intact family had 30% more pregnancies than those who did (IRR = 1.30), 

#### **Contingency Tables**

**Distribution of Marital Status by Education Level:**

```{r}
ct <- table(na.omit(data[, c("FMARITAL", "HIEDUC")]))
ct
```

**Chi-Square Test**

```{r}
chi <- chisq.test(ct)
chi
```
 
The Pearson’s Chi-squared test of independence indicated a statistically significant association between marital status and highest educational level. (p < 2.2e-16). This suggests that the distribution of education levels differs significantly across marital status categories.

**Heatmap**

```{r}
library(ggplot2)

ct_df <- as.data.frame(ct)
names(ct_df) <- c("MaritalStatus", "Education", "Count")

ggplot(ct_df, aes(x = Education, y = MaritalStatus, fill = Count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Heatmap of Marital Status by Education Level") +
  theme_minimal()
```

#### **Categorical Response Modeling**

**Ordinal Logistic Regression (HIEDUC)**

```{r}
ord_model <- polr(HIEDUC ~ AGER + HISPRACE2 + INTCTFAM + EDUCMOM, data = data, Hess = TRUE)

summary(ord_model)
```

**Odd Ratios**
```{r}

exp(coef(ord_model))

```

An ordinal logistic regression revealed that age was positively associated with higher educational attainment (OR = 1.09, p < .001). Compared to the hispanics, White individuals (OR = 1.52, p < .001) and those of Other races (OR = 2.28, p < .001) had higher odds of achieving higher education levels. Coming from a non-intact family significantly reduced the odds of attaining higher education (OR = 0.54, p < .001). While maternal education showed  a complex relationship, with both significant positive and negative coefficients.

**Multinomial Logistic Regression (FMARITAL)**


```{r}
library(nnet)


multi_model <- multinom(FMARITAL ~ AGER + HIEDUC + HISPRACE2 + INTCTFAM + EDUCMOM, data = data)
summary(multi_model)
```

**Odds-Ratios**

```{r}
coefs <- summary(multi_model)$coefficients
se <- summary(multi_model)$standard.errors

zvals <- coefs / se
pvals <- 2 * (1 - pnorm(abs(zvals)))

multi_summary <- data.frame(
  Outcome = rep(rownames(coefs), each = ncol(coefs)),
  Predictor = rep(colnames(coefs), times = nrow(coefs)),
  OR = round(exp(c(coefs)), 3),
  p = round(c(pvals), 4)
)

multi_summary
```

A multinomial logistic regression was used to examine how factors like age, education (both respondent’s and mother’s), race/ethnicity and family background affect marital status. The results showed that older age increases the chances of being widowed — for every additional year, the odds go up by 15% (OR = 1.15, p < .001). Black individuals were about 3.6 times more likely to be widowed (OR = 3.59, p = .009) compared to Hispanic individuals.

When it comes to being separated, White respondents were 3.6 times more likely (OR = 3.59, p = .050) and Black respondents were 1.9 times more likely (OR = 1.90, p = .005) than Hispanics. Additionally, people from non-intact families were over 4 times more likely to be separated (OR = 4.21, p < .001).

For those never married, each year of age slightly decreased the odds (OR = 0.88, p = .033). White individuals were much less likely to be never married (OR = 0.042, p < .001) compared to Hispanics. Higher maternal education also reduced the chances of never marrying — for example, a one-unit increase in a specific measure of maternal education cut the odds by about 67% (OR = 0.33, p = .011).

**Model Fit**

```{r}
library(pscl)

pR2(ord_model)
pR2(multi_model)
```

We assessed the model fit for both the ordinal and multinomial logistic regression models using pseudo-R² statistics. For the ordinal logistic regression model, the McFadden's R² was 0.083. The model explained approximately 29.5% (r2ML) to 30.0% (r2CU) of the variation in the outcome.

In comparison, the multinomial logistic regression model showed a better fit, with a McFadden's R² of 0.218. This model explained about 34.8% (r2ML) to 40.5% (r2CU) of the variance in the outcome. 

#### **EDA and Multiple Comparison**

**Charts**

```{r}
ggplot(data, aes(x = PREGNUM)) +
  geom_histogram(binwidth = 1, fill = "darkred", color = "black") +
  labs(title = "Histogram of Total Pregnancies", x = "Number of Pregnancies", y = "Frequency") +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = FMARITAL)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Marital Status Distribution", x = "Marital Status", y = "Count") +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = HISPRACE2)) +
  geom_bar(fill = "yellow") +
    labs(title = "Race Distribution", x = "Race", y = "Count") +
  theme_minimal()
```

```{r}
data$HIEDUC_f <- factor(data$HIEDUC,
                        levels = 1:11,
                        labels = c("Less than 12th Grade",
                                   "12th Grade, No Diploma",
                                   "GED or Equivalent",
                                   "High School Graduate",
                                   "Some College, No Degree",
                                   "Assoc. Degree (Voc)",
                                   "Assoc. Degree (Acad)",
                                   "Bachelor’s Degree",
                                   "Master’s Degree",
                                   "Professional Degree",
                                   "Doctoral Degree"))


ggplot(data, aes(x = HIEDUC_f)) +
  geom_bar(fill = "darkviolet") +
  labs(title = "Education Level Distribution",
       x = "Education Level",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}

ggplot(data, aes(x = INTCTFAM)) +
  geom_bar(fill="darkblue") +
  labs(title = "Intact vs. Non-Intact Families Distribution",
       x = "Family Structure",
       y = "Count") +
  theme_minimal()

```

```{r}

ggplot(data, aes(x = INTCTFAM, y = PREGNUM, fill = INTCTFAM)) +
  geom_boxplot() +
  labs(title = "Pregnancies by Family Structure",
       x = "Family Structure",
       y = "Number of Pregnancies",
       fill = "Family") +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = HISPRACE2, y = PREGNUM, fill = HISPRACE2)) +
  geom_boxplot() +
  labs(title = "Pregnancies by Race", x = "Race", y = "Total Pregnancies", fill = "Race") +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = HIEDUC_f, fill = HISPRACE2)) +
  geom_bar(position = "fill") + 
  labs(title = "Education Level by Race",
       x = "Education Level",
       y = "Proportion",
       fill = "Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Multiple Tests**

```{r}
kruskal_test <- kruskal.test(PREGNUM ~ HISPRACE2, data = data)
kruskal_test
```

```{r}
pairwise_result_fdr <- pairwise.wilcox.test(data$PREGNUM, data$HISPRACE2, p.adjust.method = "fdr")
pairwise_result_fdr

```

We compared the total number of pregnancies among different race groups using a Kruskal-Wallis test, which showed a significant difference (χ² = 38.54, p < 0.001). To find out which of these groups differed, we did pairwise Wilcoxon tests with False Discovery Rate (FDR) adjustments to reduce false positives. Significant differences were found between Hispanic and Other (p = 0.00022), White and Black (p = 0.00001), Black and Other (p < 0.00001), and Hispanic and Black (p = 0.01322). 